On The Subspace Learning for Network Attack Detection

### 2020-04-28 docs/thesis/chapters/2_eigen_similarity/eigen_similarity.tex
updating thesis latex

\chapter{Eigensim - Model Order Selection and Eigen Similarity based Framework for Detection and Identification of Network Attacks}
\label{ch:2_mos_eig_sim}

\begin{quotation}[]{Lao Tzu}
All difficult things have their origin in that which is easy, and great things in that which is small. 
\end{quotation}

Traditionally, cyber defense methods can be effective against ordinary and conventional types of attacks, yet may fail against innovative malicious techniques \cite{lakhina2005mining}. In order to be able to detect and avoid novel attacks and their variations, it is necessary to develop or improve techniques to achieve efficiency on resource consumption, processing capacity and response time. Moreover, it is crucial to obtain high detection accuracy and capacity to detect variations of malicious patterns. Recently, signal processing schemes have being applied to the detection of malicious traffic in computer networks \cite{Lu2009,Huang2009,Zonglin2009,david2011blind,da2012improved,tenorio2013greatest}, showing advances in network traffic analysis.

Information security may consist of both technical and procedural aspects. The former includes equipment and security systems, while the latter corresponds to security rules and recommendations. Intrusion detection and intrusion prevention systems are security systems used, respectively, to detect (passively) and prevent (proactively) threats to computer systems and computer networks. Such systems can work in the following fashions: signature-based, anomaly-based or hybrid \cite{Huang2009,mudzingwa2012study}. Additionally, anomaly detection techniques can be categorized in classification, statistical, information theory and clustering based, according to \cite{bhuyan2014network,ahmed2016survey,osanaiye2016distributed}.

This chapter proposes the Eigensim, which is an approach based on signal processing concepts applied to detection of malicious traffic in computer networks. Inspired by \cite{david2011blind,da2012improved}, the Eigensim models the network traffic using a signal processing formulation as a composition of three components: signal, artifact and noise, taking into account the incoming and outgoing traffic in certain types of network ports (TCP or UDP). The Eigensim is based on eigenvalue analysis, model order selection (MOS) and similarity analysis between eigenvectors. In contrast to \cite{david2011blind,da2012improved,tenorio2013greatest}, MOS and eigenvalue analysis are applied to detect time frames under attack. In addition, we evaluate the accuracy and performance of the proposed framework applied to a experimental scenario and to the DARPA 1998 data set \citep{osanaiye2016distributed}, which is a well known network traffic data set.

The performed experiments show that synflood, fraggle and port scan attacks can be detected accurately and with great detail in an automatic and blind fashion, i.e. in an unsupervised approach, applying signal processing concepts for traffic modeling and through approaches based on MOS and eigen similarity analysis. The main contributions of the proposed framework are the capability to blindly detect time frames under network attack via MOS and eigen analysis, and the detailed identification of the network attack via eigen similarity analysis.

This chapter is organized as follows. In Section \ref{sec:2_relatedworks}, related works are discussed. Section \ref{sec:2_datamodel} presents the data model and the evaluated data sets. Section \ref{sec:2_prop_getv} describes the proposed framework for blind and automatic detection of flood and probe attacks. Section \ref{sec:2_experimentalresults} discusses the experimental validation and presents the results, and Section \ref{sec:2_Complexity} discusses the computational complexity of the proposed framework and evaluates the required processing time for tested scenarios. Section \ref{sec:2_conclusionandfutureworks} draws the conclusions and the suggestions for future work. The Appendix \ref{apx:a_mos} presents mathematical concepts of examples of state-of-the-art MOS schemes.

@@

Data mining is often used to describe the process of extracting useful information from large databases. Multiple methods of data mining are used in \cite{he2008applying,osanaiye2016distributed} to analyze data flow in a network with the aim of identifying characteristics of malicious traffic in large scale environments. Researchers have applied data mining techniques in log analysis \cite{ghourabi2010data} to improve intrusion detection performance. However, data mining techniques used so far in network analysis require prior collection of large data sets, which is a limitation of several schemes for online analysis \cite{Huang2009}.

@@

Callegari \emph{et al} \cite{Zonglin2009} proposed a PCA-based method for identifying the traffic flows responsible for an anomaly detected at the aggregate level and evaluated their proposal through a data set with synthetic anomalies added in the data. However, Callegari \emph{et al} focus on flood attack detection, not addressing probe attack detection, and their approach relies on visual analysis.

Lee \emph{et al.} \cite{Lee2013} presented the OverSampling PCA (osPCA), which allows one to determine the anomaly of the target instance according to the variation of the resulting dominant eigenvector obtained by similarity analysis and over sampling. In contrast to Lee \emph{et al.}, the Eigensim applies MOS for detection of time frames under attack and similarity analysis to extract details for detection of time and ports under attack. Additionally, Lee \emph{et al.} only evaluate their proposed scheme for covariance analysis, while we adopt an analysis based on sample covariance of zero mean variables and sample covariance of zero mean and unitary standard deviation variables, for flood and probe attacks, respectively.

Signal processing techniques have been successfully applied to network anomaly detection \cite{Lu2009}. Lu and Ghorbani \cite{Lu2009} proposed a network anomaly detection model based on network flow, wavelet approximation, and system identification theory. However, their work requires a training method to produce a prediction model for normal daily traffic and presents limitations on identification of behaviors without significant outliers, such as port scan attacks. Zonglin \emph{et al} \cite{Zonglin2009} proposed a signal processing method to detect traffic anomaly with correlation analysis, where the correlation between traffic signals and the predicted traffic signals are used to reveal anomalies. Zonglin \emph{et al.} \cite{Zonglin2009} evaluated the correlation analysis for anomaly detection, but the work is not applied to probe and flood attack detection, simultaneously.

The data collected from honeypot systems, such as captured traffic and operating system logs, can be analyzed to obtain information about attack techniques, general trends of threats and exploits. Blind automatic detection of malicious traffic techniques have been developed for honeypots in \cite{david2011blind,da2012improved}. However, traffic on honeypot is simpler than real network traffic, because there are no running legitimate applications, i.e. background traffic, due to the fact that honeypots emulate behavior of a host within a network to deceive and lure attackers \cite{zakaria2012review}. Since honeypots do not generate legitimate traffic, the amount of data captured in honeypots is significantly lower in comparison to a Network Intrusion Detection System (NIDS), which captures and analyzes the largest possible amount of network traffic \cite{david2011blind}. MOS for blind identification of malicious activities in honeypots was proposed by us in \cite{david2011blind}, which evaluated criteria for selecting the model order, through simulations and comparing the order of the resulting model with the true model order.

The proposed framework does not require either a significant amount of logs to detect attacks, nor prior data collection, in order to make comparisons and evaluate the existence of malicious traffic. The proposed solution is automatic and blind for detection of time frames under probe and flood attacks through MOS and eigen analysis. Moreover, we apply eigen similarity analysis to identify details of time and ports under network attacks.

Several approaches for network attack detection uses the KDD 99 \cite{ji2016multi,ahmed2016survey,osanaiye2016distributed,bhuyan2014network} data sets for accuracy and performance evaluation, due to their availability and labeled attacks. Even though the KDD 99 data set are criticized by for the generation procedure and the risk of over-estimations of anomaly detection due to data redundancy, it still represents one of the few publicly available labeled data sets adopted by researchers \cite{osanaiye2016distributed,bhuyan2014network}.

NSL-KDD \cite{tavallaee2009detailed} data set is the refined version of the KDD 99 data set that redundant data records are removed, in order to avoid biased classifications. Additionally, some approaches uses simulated \cite{callegari2011novel} scenarios or non-public data sets their evaluations. Since the proposed approach relies on a packet level analysis and the KDD 99 and NSL-KDD data sets adopt a traffic aggregation by connections, we consider the use of a experimental scenario on a real network and the DARPA 1998 data set, which is the source for the creation of the KDD 99 and NSL-KDD datas ets. Note that the proposed approach is not based on learning or classification techniques, which are more susceptible to biased results caused by the issues in the DARPA/KDD data sets.

@@

In this work, scalars are denoted by italic letters ($a, b, A, B, \alpha, \beta$), vectors by lowercase bold letters ($\pmb{a}, \pmb{b}$), matrices by uppercase bold letters ($\pmb{A}, \pmb{B}$), and $a_i,_j$ denotes the (\emph{i, j}) elements of the matrix $\pmb{A}$. The superscripts \textsuperscript{T} and \textsuperscript{-1} are used for matrix transposition and matrix inversion, respectively. We define the operator $\rm diag(\cdot)$ that returns the vector of the main diagonal of a given matrix, the operator $\rightarrow$, which denotes the deletion of a given element from a set and the operator $\#$, that returns the rank of a matrix, and the operator $\sim$ that sorts the elements of a vector in ascending order.

This section also presents details of the experimental scenario and the selected cases of the DARPA data set, along with a description of the data set model as a signal superposition of signal, artifact and noise, that refer to legitimate, control and malicious traffic. Subsection \ref{sec:2_DataCollection} describes the environment and scenario adopted in order to reproduce flood and probe attacks. Subsection \ref{sec:2_ModelingData} presents how network traffic can be modeled as signal superposition. Subsection \ref{sec:2_SynfloodFraggleandPortscan} details the traffic of synflood, fraggle and port scan attacks, and Subsection \ref{sec:2_Darpadata set} discusses the use of the DARPA data set for evaluation of the proposed approach.

@@

%TODO - recreate the figure and change the concepts
\begin{figure}[h!]
     \centering 
     \includegraphics[width=11cm]{figures/ch2/fig09.png}
     \caption{Scenario to reproduce legitimate traffic, noise, flood and port scan.}
     \label{fig:2.01}
\end{figure}

Note that the set of network traffic is modeled as legitimate, controls and malicious traffic, which refer to signal, artifact and noise, respectively. In many organizations the web based traffic is predominant, since most of corporate services are  web pages, customized web-based systems and cloud services. It is possible to characterize the traffic of a DHCP service as an example of artifact associated with the transport layer, as well as it would be possible to classify heart beat traffic and other seasonal and controlled traffic as artifact. For malicious traffic, three types of networks attacks are evaluated: synflood, fraggle and port scan. Here we refer to port scan as an attack, according to \cite{ahmed2016survey,moustafa2019holistic}, however it is one approach usually adopted for acquire information in order to perform an attack. These attacks are reproduced using well-known security tools, such as Nmap\footnote{http://nmap.org} for port scan, Metasploit\footnote{http://www.metasploit.com} for synflood and Hping\footnote{http://hping.org} to lead the fraggle attack.

@@

In the proposed framework, the goal is to detect the anomalies only taking into account the traffic profile, i.e. specific information such as origin or destination IP, behavioral pattern or content of the attack are not considered. Therefore, IP spoofing or data encryption would not cause impact to the proposed approach and evaluation, since our proposal only relies on the timestamp (for sequencing) and port number.

@@

By modeling the data set as a signal superposition, the network traffic ($\pmb{X}$) can be characterized as a mixture of three components: legitimate traffic ($\pmb{U}$), artifact ($\pmb{A}$) and malicious traffic ($\pmb{N}$), according to the following expression:

\begin{equation}\label{eq:2.01}
	\pmb{X}^{(q)} = \pmb{U}^{(q)} + \pmb{A}^{(q)} + \pmb{N}^{(q)},
\end{equation}
where $q$ represents the $q$-th time frame, which is a time aggregation of network traffic. The matrix $\pmb{X}^{(q)} \in \mathbb{R}^{M \times N}$ consists of \emph{M} rows and \emph{N} columns, where each row represents a communication port, and each column represents time bins of a defined size, such as one minute. Each element $x_{m,n}^{(q)}$ stands for the number packets that appears at $n$-th minute for the port $m$, during the $q$-th time frame.

The legitimate traffic $\pmb{U}^{(q)}$ is characterized by the traffic from user's ordinary operations, such as when performing access to web pages, as well when there is the traffic required to domain name resolution. Figure \ref{fig:2.03} depicts an example of the legitimate traffic simulated during the experiments.

% TODO - Replace
\begin{figure}[h!]
     \centering 
     \includegraphics[width=11cm]{figures/ch2/fig03.png}
     \caption{Traffic from user's operations, that can be characterized by web access, traffic of well-known applications or network protocols.}
     \label{fig:2.03}
\end{figure}

The traffic that is not associated with user's operations and with malicious traffic is modeled as artifact $\pmb{A}^{(q)}$. The acquisition of logical IP network address (DHCP) is an example of artifact, where independently of any user operation, the machine receives an IP address, since it is configured to automatically perform a DHCP address request. Figure \ref{fig:2.04} depicts an example of artifact in a network traffic, by means of traffic to ports 67 and 68. The traffic coming from a malicious activity, i.e. port scanning, synflood or fraggle attacks, is represented by the matrix $\pmb{N}^{(q)}$.

% TODO - Replace
\begin{figure}[h!]
     \centering 
     \includegraphics[width=11cm]{figures/ch2/fig04.png}
     \caption{Network traffic of user independent operations for network management.}
     \label{fig:2.04}
\end{figure}

We define that if the $\#\pmb{X}^{(q)} ≠ 0$, then there is malicious traffic in the evaluated time frame $q$, on the other hand, if the $\#\pmb{X}^{(q)} = 0$, then there is no malicious traffic. We show how to detect the $\#\pmb{X}^{(q)}$, given only the matrix $\pmb{X}^{(q)}$, in order to identify malicious traffic.

@@

A DoS is an attempt by an attacker to prevent legitimate access to websites by overwhelming the amount of available bandwidth or resources of the computer system. DoS is implemented by either forcing targets to be unavailable through the exploiting of system vulnerabilities, or consuming resources through large amount of network traffic, characterizing flood attacks.

Probe attacks scan computer and network systems to collect information about the host, such as open ports, topology, running software or version of technologies, in order to find vulnerabilities.

@@

% TODO - Replace
\begin{figure}[h!]
     \centering 
     \includegraphics[width=11cm]{figures/ch2/fig05.png}
     \caption{A large quantity of SYN requests to a target, in order to cause a DoS.}
     \label{fig:2.05}
\end{figure}

Regarding the fraggle attack, large packets with UDP echo segments are sent to the broadcast address of a network. Every packet is modified to have the source address of the victim, in order to implement the source address spoofing technique. Therefore, each host receives a huge amount of requests UDP echo and all of them replies to the IP address of the victim, causing a packet flooding aiming a DoS. Figure \ref{fig:2.06} depicts an example of the fraggle attack in a real computer network.

The fraggle attack can affect the entire network, since all hosts receive several requests UDP echo and respond with the ICMP protocol, therefore each host acts as an amplifier of the attack. This part of the fraggle attack is not taken into account in this work, because the victim receives ICMP (network layer) packets originated from the hosts that are attacked with flooding packet UDP echo.

% TODO - Replace
\begin{figure}[h!]
     \centering 
     \includegraphics[width=11cm]{figures/ch2/fig06.png}
     \caption{Large amount of “UDP echo” requests and replies, causing packet flooding.}
     \label{fig:2.06}
\end{figure}

The Figure \ref{fig:2.06} shows that more than 6,000,000 malicious packets can be counted in an interval of ten minutes, which can be considered an abnormal network traffic, especially due to the concentrated traffic in a short period of time and due to the similarity of the outstanding traffic.

@@

\begin{figure}[h!]
     \centering 
     \includegraphics[width=11cm]{figures/ch2/fig07.png}
     \caption{Connection attempts in order to identify active ports.}
     \label{fig:2.07}
\end{figure}

Figure \ref{fig:2.07} depicts an example of the port scan attack in a real computer network. Note that the traffic is composed of two packets for each TCP port and one UDP packet to each port. The incoming and outgoing packets analysis, for each port, shows the high correlation and similarity of TCP and UDP traffic during the simulated port scan attack.

@@

We observe that the most cases of DoS of DARPA 98 focus on exploit system vulnerabilities instead of on flooding attack. One example is the occurrence of a neptune attack which sends 20 SYN packets, what is a behavior that differs of the expected flooding attack behavior. Therefore, there were selected the cases that simulates several network traffic or numerous connection requests, also known as flooding attack \cite{ahmed2016survey,osanaiye2016distributed}, and the cases that scan ports sending just a few packets. From the simulated probe attacks, we select the cases that rely on TCP or UDP connections.

@@

This section describes the proposed technique to detect synflood, fraggle and port scan, according to the overview depicted in Figure \ref{fig:2.08}, which represents the proposed framework for detection and identification of network attacks. In Subsection \ref{sec:2_prop_LargestEigenvaluebyTimeFrames} we present the steps for extraction of the largest eigenvalue for each $q$-th time frame. Next, in Subsection \ref{sec:2_prop_MOSSchemes}, we show how to apply the eigenvalues on the MOS scheme in order to detect the attack. In Subsection \ref{sec:2_prop_EigenvalueAnalysis}, we present the eigenvalue analysis to identify the time frames detected as under attack, and the Subsection \ref{sec:2_prop_EigenSimilarityAnalysis} describes the similarity analysis evaluated for detailed attack identification.

\begin{figure}[h!]
	\centering
     \includegraphics[width=11cm]{figures/ch2/mos_eigen_similarity.eps}
     \caption{Overview of The Framework for Detection and Identification of Network Attacks.}
     \label{fig:2.08}
\end{figure}

@@

The proposed attack detection algorithm starts by the data pre-processing of a network traffic log containing IP, ports and timestamp of senders and receivers. During this step, the desired information is extracted in order to classify and to count packets according to the origin and destination ports. Subsequently, this information is grouped by minutes and by time frames.

With the data grouped into $Q$ time frames, the framework considers the time variations of the matrix $\pmb{X}^{(q)} \in \mathbb{R}^{M\times{N}}$, with $q = 1, \ldots, Q$, in order to detect the attack.

>>>>

\chapter{Eigensimilarity based Framework for Detection and Identification of Network Attacks}
\label{ch:2_mos_eig_sim}

According to \cite{Hevesi2019}, at times, organizations relate their security measures to confidentiality and integrity, ignoring availability. However, according to \cite{BISSELL2019}, the global average annual cost of DoS attack was US\$1.7 million in 2018.

The Denial of Service (DoS) attack attempts to deny access to system or network resources, attacking the availability of the service, by means of flooding techniques for consuming the available resources, or by means of subtle approaches which send small amount of data that can cause failure and unavailability of the system. In 2017 the top motivation behind DDoS attacks was criminals demonstrating attack capabilities, with gaming and criminal extortion attempts in second and third place, respectively \cite{cisco2019}.

With application layer Distributed Denial of Service (DDoS) attacks continuing to rise, vendors have begun adding DDoS mitigation features into solutions that protect web applications \cite{Hevesi2019}. The leading solutions for monitoring and mitigation of DDoS attacks in application layer start with behavioral analytics and supplement with signature-based detection for known malicious application attacks \cite{Hevesi2019}.

Probe attacks aim to scan the networks to identify running services, open ports, running services and vulnerabilities that can be exploited. A probe attack is considered the first step in an attack to compromise a host or network. Although no specific damage is caused by these attacks, they are considered serious threats by \cite{ahmed2016survey} and \cite{moustafa2019holistic}. Note that we refer to port scan as an attack, according to \cite{ahmed2016survey}.

Anomaly-based systems for network attack detection focus on finding exceptional, suspicious or rare observations in network traffic that do not conform to the expected normal behavior \cite{bhuyan2014network}. It is possible to adopt unsupervised or semi-supervised approaches for network anomaly detection, by means of behavioral analysis, where known anomalies for training models are not necessary \cite{moustafa2019holistic}.

Recently, signal processing schemes have been applied for network anomaly detection in computer networks by means of unsupervised approaches, showing advances in network traffic analysis \cite{Zonglin2009}, in network attack detection \cite{callegari2011novel}, and in network anomaly detection based on network flows \cite{Lu2009}.

Considering that a subspace learned by eigenvalue decomposition can highlight anomalies for network attack detection by means of model order selection (MOS), and that the comparison between the principal eigenvectors can reveal anomalies when comparing legitimate and anomalous observations. In this chapter we propose the Eigensimilarity, which is an approach based on signal processing concepts applied to detection of malicious traffic in computer networks, with focus on DoS and portscan attacks.

The Eigensimilarity is based on eigenvalue analysis, model order selection (MOS) and similarity analysis between the principal eigenvectors. In contrast to \cite{david2011blind, da2012improved, tenorio2013greatest}, MOS and eigenvalue analysis are applied to detect time frames under attack, and the analysis of similarity between the principal eigenvectors are used for precise identification of the time and ports under attack. In addition, we evaluate the accuracy and performance of the proposed framework applied to an experimental scenario and to the DARPA 1998 data set \citep{osanaiye2016distributed}, which is a well-known network traffic data set.

The performed experiments show that synflood, fraggle and port scan attacks can be accurately detected and with great detail in an automatic and blind fashion, i.e. in an unsupervised approach that does not require data for training, applying signal processing concepts for traffic modeling and through approaches based on MOS and principal eigenvector similarity analysis. The main contributions of the proposed framework are the capability to blindly detect time frames under network attack via MOS and eigen analysis, and the detailed identification of the network attack via principal eigenvector similarity analysis.

This chapter is organized as follows. In Section \ref{sec:2_relatedworks} the related works are discussed. We present the data model and the evaluated data sets in Section \ref{sec:2_datamodel} and the proposed framework for blind and automatic detection of flood and probe attacks is described in Section \ref{sec:2_prop_getv}. The discussion of the experimental validation and results are presented in Section \ref{sec:2_experimentalresults}, while the discussion of the computational complexity and evaluation of the required processing time for tested scenarios are presented in Section \ref{sec:2_Complexity}. In Section \ref{sec:2_conclusionandfutureworks} we draw the conclusions and the suggestions for future work.

@@

Data mining is often used to describe the process of extracting useful information from large databases. Multiple methods of data mining are used in \cite{he2008applying,osanaiye2016distributed} to analyze data flow in a network with the aim of identifying characteristics of malicious traffic in large scale environments. Researchers have applied data mining techniques in log analysis \cite{ghourabi2010data} to improve intrusion detection performance. However, data mining techniques used so far in network analysis require prior collection of large data sets, which is a limitation of several schemes for online analysis \cite{he2008applying}.

@@

Callegari \emph{et al.} \cite{callegari2011novel} proposed a PCA-based method for identifying the traffic flows responsible for an anomaly detected at the aggregate level and evaluated their proposal through a data set with synthetic anomalies added in the data. However, Callegari \emph{et al.} \cite{callegari2011novel} focus on flood attack detection, not addressing probe attack detection, and their approach relies on visual analysis.

Lee \emph{et al.} \cite{Lee2013} presented the OverSampling PCA (osPCA), which allows one to determine the anomaly of the target instance according to the variation of the resulting dominant eigenvector obtained by similarity analysis and over sampling. In contrast to Lee \emph{et al.} \cite{Lee2013}, the Eigensimilarity applies MOS for detection of time frames under attack and similarity analysis to extract details for detection of time and ports under attack. Additionally, Lee \emph{et al.} \cite{Lee2013} only evaluate their proposed scheme for covariance analysis, while we adopt an analysis based on sample covariance of zero mean variables and sample covariance of zero mean and unitary standard deviation variables, for flood and probe attacks, respectively.

Signal processing techniques have been successfully applied to network anomaly detection \cite{Lu2009}. Lu and Ghorbani \cite{Lu2009} proposed a network anomaly detection model based on network flow, wavelet approximation, and system identification theory. However, their work requires a training method to produce a prediction model for normal daily traffic and presents limitations on identification of behaviors without significant outliers, such as port scan attacks. Zonglin \emph{et al.} \cite{Zonglin2009} proposed a signal processing method to detect traffic anomaly with correlation analysis, where the correlation between traffic signals and the predicted traffic signals are used to reveal anomalies. Zonglin \emph{et al.} \cite{Zonglin2009} evaluated the correlation analysis for anomaly detection, but the work is not applied for probe attack detection and does not evaluate results for attack detection on DARPA data set.

The data collected from honeypot systems, such as captured traffic and operating system logs, can be analyzed to obtain information about attack techniques, general trends of threats and exploits \cite{david2011parallel}. Blind automatic detection of malicious traffic techniques have been developed for honeypots in \cite{david2011blind, da2012improved}. However, traffic on honeypot is simpler than real network traffic, because there are no running legitimate applications, i.e. background traffic, due to the fact that honeypots emulate behavior of a host within a network to deceive and lure attackers \cite{zakaria2012review}. Since honeypots do not generate legitimate traffic, the amount of data captured in honeypots is significantly lower in comparison to a Network Intrusion Detection System (NIDS), which captures and analyzes the largest possible amount of network traffic \cite{david2011blind}.

MOS for blind identification of malicious activities in honeypots was proposed by us in \cite{david2011blind}, which evaluated criteria for selecting the model order, through simulations and comparing the order of the resulting model with the true model order.

The proposed framework does not require either a significant amount of logs to detect attacks, nor prior data collection, in order to make comparisons and evaluate the existence of malicious traffic. The proposed solution is automatic and blind for detection of time frames under probe and flood attacks through MOS and eigen analysis. Moreover, we apply principal eigenvector similarity analysis to identify details of time and ports under network attacks.

Ahmed \emph{et al.} \cite{ahmed2016survey} highlight a observed lack of publicly available labeled data set for network anomaly detection, and discuss recent research related to overcome the issues of publicly available network intrusion evaluation data sets, including a list of current data sets and an evaluation of attack types and available labels. According to \cite{ahmed2016survey}, the available data sets present limitations on availability of labeled traffic for classification of background, legitimate and malicious traffic. The data imbalance between legitimate and malicious traffic is a important concern for real world problems of network attack, but the principal data sets usually adopt a balanced distribution between legitimate and malicious traffic, what make the attack detection more feasible for general classifiers and machine learning algorithms than for anomaly detection algorithms. Moreover, according to Ahmed \emph{et al.} \cite{ahmed2016survey}, the most adopted data set are not up-to-date with current or novel attacks, tools, operational systems and network topology.

Several approaches for network attack detection uses the KDD 99 \cite{ji2016multi, ahmed2016survey, osanaiye2016distributed, bhuyan2014network} data sets for accuracy and performance evaluation, due to their availability and labeled attacks. Even though the KDD 99 data set are criticized by for the generation procedure and the risk of over-estimations of anomaly detection due to data redundancy, it still represents one of the few publicly available labeled data sets adopted by researchers \cite{osanaiye2016distributed, bhuyan2014network}. NSL-KDD \cite{tavallaee2009detailed} data set is the refined version of the KDD 99 data set that redundant data records are removed, in order to avoid biased classifications. Additionally, some approaches uses simulated \cite{callegari2011novel} scenarios or non-public data sets their evaluations.

Since the proposed approach relies on a packet level analysis and the KDD 99 and NSL-KDD data sets adopt a traffic aggregation by connections, we consider the use of an experimental scenario on a real network and the DARPA 1998 data set, which is the source for the creation of the KDD 99 and NSL-KDD data sets. Note that the proposed approach is not based on learning or classification techniques, which are more susceptible to biased results caused by the issues in the DARPA/KDD data sets.

@@

In this work, scalars are denoted by italic letters ($a, b, A, B, \alpha, \beta$), vectors by lowercase bold letters ($\pmb{a}, \pmb{b}$), matrices by uppercase bold letters ($\pmb{A}, \pmb{B}$), and $a_i,_j$ denotes the (\emph{i, j}) elements of the matrix $\pmb{A}$. The superscripts \textsuperscript{T} and \textsuperscript{-1} are used for matrix transposition and matrix inversion, respectively. The $l_2$ norm is denoted by $\norm{\cdot}$. We define the operator $\rm diag(\cdot)$ that returns the vector of the main diagonal of a given matrix, the operator $\rightarrow$, which denotes the deletion of a given element from a set and the operator $\#$, that returns the rank of a matrix, and the operator $\sim$ that sorts the elements of a vector in ascending order.

This section also presents details of the experimental scenario and the selected cases of the DARPA data set. In Subsection \ref{sec:2_DataCollection} we describe the environment and scenario adopted in order to reproduce flood and probe attacks. In Subsection \ref{sec:2_ModelingData} are presented how network traffic can be modeled as signal superposition. We detail, in Subsection \ref{sec:2_SynfloodFraggleandPortscan}, the traffic of synflood, fraggle and port scan attacks, and in Subsection \ref{sec:2_Darpadata set} are discussed the use of the DARPA data set for evaluation of the proposed approach.

@@

\begin{figure}[h!]
     \centering 
     \includegraphics[width=9cm]{figures/ch2/scenario.eps}
     \caption{Scenario to reproduce legitimate traffic, flood and port scan.}
     \label{fig:2.01}
\end{figure}

In many organizations the web based traffic is predominant, since most of corporate services are  web pages, customized web-based systems and cloud services. It is possible to characterize the traffic of a Dynamic Host Configuration Protocol (DHCP) service as an example of legitimate associated with the application layer, as well as it would be possible to classify seasonal and controlled traffic as legitimate traffic. For malicious traffic, three types of networks attacks are evaluated: synflood, fraggle and port scan. Here we refer to port scan as an attack, according to \cite{ahmed2016survey, moustafa2019holistic}, however it is one approach usually adopted for acquire information in order to perform an attack. These attacks are reproduced using well-known security tools, such as Nmap\footnote{http://nmap.org} for port scan, Metasploit\footnote{http://www.metasploit.com} for synflood and Hping\footnote{http://hping.org} to lead the fraggle attack.

@@

In the proposed framework, the goal is to detect the anomalies only taking into account the traffic profile, i.e. specific information such as origin or destination IP, behavioral pattern or content of the attack are not considered. Therefore, IP spoofing or data encryption would not cause impact to the proposed approach and evaluation, since our proposal only relies on the timestamp (for sequencing) and destination port number.

@@

By modeling the data set as a signal superposition, the network traffic ($\pmb{X}$) can be characterized as a mixture of two components: legitimate traffic ($\pmb{U}$) and malicious traffic ($\pmb{N}$), according to the following expression:

\begin{equation}\label{eq:2.01}
	\pmb{X}^{(q)} = \pmb{U}^{(q)} + \pmb{N}^{(q)},
\end{equation}
where $q$ represents the $q$-th time frame, which is a time aggregation of network traffic. The matrix $\pmb{X}^{(q)} \in \mathbb{R}^{M \times N}$ consists of \emph{M} rows and \emph{N} columns, where each row represents a communication port, and each column represents time bins of a defined size, such as one minute. Each element $x_{m,n}^{(q)}$ stands for the number packets that appears at $n$-th minute for the port $m$, during the $q$-th time frame.

The legitimate traffic $\pmb{U}^{(q)}$ is characterized by user's ordinary operations and by legitimate traffic that are automatically generated for network management and for background services. The access to web pages by users or the name resolution by means of the Domain Name System (DNS) are examples of legitimate traffic generated by user operations, as can be seen in Figure \ref{fig:2.03}, which depicts the legitimate traffic simulated to reproduce user's operations during the experiments.

\begin{figure}[h!]
     \centering 
     \includegraphics[width=11cm]{figures/ch2/user_traffic.eps}
     \caption{Traffic from user's operations.}
     \label{fig:2.03}
\end{figure}

The Figure \ref{fig:2.04} depicts an example of legitimate traffic of user independent operations, by means of traffic to ports 67 and 68, where it is possible to observe a low amount of packets. The acquisition of logical IP network address by means of DHCP is an example of legitimate traffic not associated with a user, where independently of any user operation, the machine receives an IP address, since it is configured to automatically perform a DHCP address request.

\begin{figure}[h!]
     \centering 
     \includegraphics[width=11cm]{figures/ch2/auto_traffic.eps}
     \caption{Network traffic of user independent operations for network management.}
     \label{fig:2.04}
\end{figure}

The traffic coming from a malicious activity, i.e. port scanning, synflood or fraggle attacks, is represented by the matrix $\pmb{N}^{(q)}$. We define that if the rank of $\pmb{X}^{(q)}$ is not zero, according to $\#\pmb{X}^{(q)} \neq 0$, which denotes that the rank of the $q$-th $\pmb{X}$ is different of 0, then there is malicious traffic in the evaluated time frame $q$. On the other hand, if the $\#\pmb{X}^{(q)} = 0$, then there is no malicious traffic in time frame $q$. We show how to detect the $\#\pmb{X}^{(q)}$, given only the matrix $\pmb{X}^{(q)}$, in order to identify malicious traffic.

@@

A DoS is an attempt by an attacker to prevent legitimate access to websites by overwhelming the amount of available bandwidth or resources of the computer system. DoS is implemented by either forcing targets to be unavailable through the exploiting of system vulnerabilities, or consuming resources through large amount of network traffic, characterizing flood attacks. Probe attacks scan computer and network systems to collect information about the host, such as open ports, topology, running software or version of technologies, in order to find vulnerabilities.

@@

\begin{figure}[h!]
     \centering 
     \includegraphics[width=11cm]{figures/ch2/synflood_traffic.eps}
     \caption{A large quantity of SYN requests to a target, in order to cause a DoS.}
     \label{fig:2.05}
\end{figure}

Regarding the fraggle attack, large packets with UDP echo segments are sent to the broadcast address of a network. Every packet is modified to have the source address of the victim, in order to implement the source address spoofing technique. Therefore, each host receives a huge amount of requests UDP echo and all of them replies to the IP address of the victim, causing a packet flooding aiming a DoS.

The Figure \ref{fig:2.06} depicts an example of the fraggle attack in a real computer network, and shows that more than 6,000,000 malicious packets can be counted in an interval of ten minutes, which can be considered an abnormal network traffic, due to the concentrated traffic in a short period of time and due to the similarity of the outstanding traffic.

The fraggle attack can affect the entire network, since all hosts receive several requests UDP echo and respond with the  Internet Control Message Protocol (ICMP), therefore each host acts as an amplifier of the attack. This part of the fraggle attack is not taken into account in this work, because the victim receives ICMP packets originated from the hosts that are attacked with flooding packet UDP echo.

\begin{figure}[h!]
     \centering 
     \includegraphics[width=11cm]{figures/ch2/fraggle_traffic.eps}
     \caption{Large amount of “UDP echo” requests and replies, causing packet flooding.}
     \label{fig:2.06}
\end{figure}

@@

\begin{figure}[h!]
     \centering 
     \includegraphics[width=11cm]{figures/ch2/portscam_traffic.eps}
     \caption{Connection attempts in order to identify active ports.}
     \label{fig:2.07}
\end{figure}

Figure \ref{fig:2.07} depicts an example of port scan attack in a real computer network. We simulated a traffic with two packets for each TCP port and one UDP packet to each port. The incoming and outgoing packets analysis, for each port, shows the high correlation and similarity of TCP and UDP traffic during the simulated port scan attack.

@@

The most cases of DoS of DARPA 98 focus on exploit system vulnerabilities instead of on flooding attack. One example is the occurrence of a neptune attack which sends 20 SYN packets, what is a behavior that differs of the expected flooding attack behavior. Therefore, there were selected the cases that simulates several network traffic or numerous connection requests, also known as flooding attack \cite{ahmed2016survey, osanaiye2016distributed}, and the cases that scan ports sending just a few packets. From the simulated probe attacks, we select the cases that rely on TCP or UDP connections.

@@

According to the overview depicted in Figure \ref{fig:2.08}, we present in this section the proposed framework for detection and identification of network attacks.

In Subsection \ref{sec:2_prop_LargestEigenvaluebyTimeFrames} we present the steps for extraction of the largest eigenvalue for each $q$-th time frame. Next, in Subsection \ref{sec:2_mos} are presented the mathematical concepts and examples of state-of-the-art MOS schemes, and how to apply the eigenvalues on the MOS scheme in order to detect the attack. In Subsection \ref{sec:2_prop_EigenvalueAnalysis}, we present the eigenvalue analysis to identify the time frames detected as under attack, and the Subsection \ref{sec:2_prop_EigensimilarityAnalysis} describes the similarity analysis evaluated for detailed attack identification.

\begin{figure}[h!]
	\centering
     \includegraphics[width=12cm]{figures/ch2/mos_eigen_similarity.eps}
     \caption{Overview of the framework for detection and identification of network attacks.}
     \label{fig:2.08}
\end{figure}

@@

The proposed attack detection algorithm starts by the data preprocessing of a network traffic log containing IP, ports and timestamp of senders and receivers. During this step, the desired information is extracted in order to count packets according to the destination ports by time. Subsequently, this information is grouped by minutes and by time frames.
